---
title: "CosMx SMI Data Analysis"
output: github_document
autor : "Jose Liboy-Lugo"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading required packages and variables

```{r load-packages}
#usethis::edit_r_environ("project") #Set the environment variables for the project.

library(Seurat)
library(dplyr)
library(ggplot2)
library(data.table)
library(ggrepel)

date <- Sys.Date()
source(here::here("src", "spatial_uproc.R"))
source(here::here("src", "spatial_qc.R"))
source(here::here("src", "spatial_dproc.R"))
                         
COSMX_INPUTFILES_REPO        <- Sys.getenv("input_path")
COSMX_OUTPUTFILES_REPO       <- Sys.getenv("output_path")
PANEL_FOV                    <- "Hs_UCC"
PANEL_PLEX                   <- 1000
NEG_PANEL_PLEX               <- 10
FOV_MAX_PROP_LOSS            <- 0.6

# QC thresholds
COUNTS_THRESHOLD_MIN          <- 20
COUNTS_THRESHOLD_MAX          <- NULL
AREA_THRESHOLD_MIN            <- NULL
AREA_THRESHOLD_MAX            <- 60e3
SBR_THRESHOLD_MIN             <- 0
SBR_THRESHOLD_MAX             <- NULL
FOV_INTEGRITY_PERCENT_THRESHOLD <- 0.6
SPLIT_RATIO_THRESHOLD_MIN     <- NULL
SPLIT_RATIO_THRESHOLD_MAX     <- log2(4)

options(future.globals.maxSize = 30 * 1024^3) #Set the maximum allowed RAM memory to x * 1024^3 = xGB

```

## Loading CosMx seurat objects

``` {r load-seurat-objects}

obj <- readCosMxSeurat(experiment_repo = COSMX_INPUTFILES_REPO)
obj

```

## Plotting FOV QC metrics to evaluate data quality before running cell QC

``` {r plotting_qc}

#Plotting mean transcript per cell per fov. 
fov_positions <- fov_plotting(COSMX_INPUTFILES_REPO, obj)

#Saving fov qc plots
for (i in seq_along(fov_positions)) {
  ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_fov_qc_plots_", names(fov_positions)[[i]], ".png"), plot = fov_positions[[i]], width = 10, height = 8, units = "in")
}

#Plotting mean transcript per cell per fov boxplots
meantranscript_per_cell_per_fov_boxplot <- fov_box_plot(obj)
ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_meantranscript_per_cell_per_fov_boxplot.png"), plot = meantranscript_per_cell_per_fov_boxplot, width = 10, height = 8, units = "in")

#Plotting mean staining intensity by slide and marker
staining_qc_plot <- staining_qc_plot(obj)
ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_staining_qc_plot.png"), plot = staining_qc_plot, width = 10, height = 8, units = "in")

#Creating a slide QC table summarizing key QC metrics for each slide. 
qc_table <- slide_qc_table(obj, PANEL_PLEX, NEG_PANEL_PLEX)
write.csv(qc_table, file = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_slide_qc_table.csv"), row.names = FALSE)


```

## Runing cell and FOV QC

```{r cell_qc}

obj <- computeFOVintegrity(obj, max_prop_loss = FOV_MAX_PROP_LOSS, panel_name = PANEL_FOV)
obj <- computeSBR(obj)
obj <- computeSplitRatio(obj)
obj$SplitRatioToLocal_log2 <- log2(obj$SplitRatioToLocal)

```

## Plotting the QC results 

```{r plotting_qc}
# Raw QC scatter plots with thresholds
plots <- plotQCs(
  obj,
  count_min       = COUNTS_THRESHOLD_MIN,
  count_max       = COUNTS_THRESHOLD_MAX,
  area_min        = AREA_THRESHOLD_MIN,
  area_max        = AREA_THRESHOLD_MAX,
  sbr_min         = SBR_THRESHOLD_MIN,
  sbr_max         = SBR_THRESHOLD_MAX,
  split_ratio_min = SPLIT_RATIO_THRESHOLD_MIN,
  split_ratio_max = SPLIT_RATIO_THRESHOLD_MAX,
  base_size = 15) %>%
  patchwork::wrap_plots(nrow = 4)

ggplot2::ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_qc_plots.png"), plot = plots, width = 10, height = 15, units = "in")

```

## FLagging cells based on the fov and cell QCs. 

Notice that I am not removing cells, just flagging. 

```{r flagging cells based on qc}

obj <- qc_flagging(
  obj,
  count_min               = COUNTS_THRESHOLD_MIN,
  count_max               = COUNTS_THRESHOLD_MAX,
  area_min                = AREA_THRESHOLD_MIN,
  area_max                = AREA_THRESHOLD_MAX,
  fov_integrity_threshold = FOV_INTEGRITY_PERCENT_THRESHOLD,
  sbr_min                 = SBR_THRESHOLD_MIN,
  sbr_max                 = SBR_THRESHOLD_MAX,
  split_ratio_min         = SPLIT_RATIO_THRESHOLD_MIN,
  split_ratio_max         = SPLIT_RATIO_THRESHOLD_MAX)

```


## Removing flagged cells for downstream analysis

In this section, I am subsetting the seurat object to only include cells with 'flag_overall' == FALSE

``` {r removing flagged cells}

#remove flagged cells
obj <- subset(obj, obj$flag_overall == FALSE)
obj

```
## Assingning patient IDs to each core

In this section, I am using the `assignCoresToCells()` function to assing patient IDs 
to each individual core. 

At the end, I removed FOVs with mislabeled cells. I figured that it is easier to remove those 
FOVs, rather than relabeling cells. 

``` {r core_assignment}

tma_map <- readxl::read_excel(path = Sys.getenv("tma_map_path"), sheet = "FT- OTHER CA TMA Map", col_names = FALSE)

fov_positions <- read.csv(file = Sys.getenv("fov_positions_path"))

tma_map <- as.matrix(tma_map)
tma_map <- matrix(sub(",.*", "", tma_map), nrow = nrow(tma_map), ncol = ncol(tma_map))
tma_map[tma_map == "" | tma_map == "NA" | tma_map == "NULL"] <- NA

tma_map <- tma_map[nrow(tma_map):1, ] # flip rows to match TmaPlot orientation



obj <- assignCoresToCells(obj = obj, tma_map = tma_map, fov_positions = fov_positions)

tma_plot <- TmaPlot(obj, col.by = "sample_id_col", dark = F)
ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_tma_sampleid.png"), plot = tma_plot, width = 8, height = 6, units = "in")

#Filter TMA to remove mislabeled FOVs. 
exclude_fov <- c(28, 29, 30, 67, 69, 70, 71, 77, 95, 108, 109, 110, 111, 113, 114, 
                  241, 242, 243, 245, 246, 247, 248, 278, 279, 280, 281, 282, 283, 
                  284, 285, 286, 331, 332, 333, 334, 335, 336, 337, 338, 339, 340)

obj <- subset(obj, !(FOV %in% exclude_fov))

```


## Normalizing data with scPearsonPCA 

Here, I am normalizing data with the scPearsonPCA package, which have proven 
quite effective with CosMx data. 

``` {r normalization with scPearsonPCA}

obj <- scpearson_norm(obj, 
                      assay = "RNA", 
                      new.pca.name = "pearsonpca", 
                      new.umap.name = "pearsonumap", 
                      var_features = NULL) 

obj <- FindNeighbors(obj, reduction = "pearsonumap", graph.name = "pearson_snn", k.param = 50, annoy.metric = "euclidean", dims = 1:2)

obj <- FindClusters(obj, graph.name = "pearson_snn", cluster.name = "pearson_leiden",
                                algorithm = 4, resolution = 0.1)  # Leiden = algorithm 4

umap_leiden = DimPlot(obj, reduction = "pearsonumap", group.by = "pearson_leiden", label = TRUE, repel = TRUE)

ggsave(filename = paste0(COSMX_OUTPUTFILES_REPO, "/", date, "_pearsonumap_leiden_tma.png"), plot = umap_leiden, width = 8, height = 6, units = "in")

```


## Normalizing data with scPearsonPCA and integrated batch correction

``` {r normalization with scPearsonPCA and batch correction}

obj <- scpearson_norm_batch(obj, 
                            assay = "RNA", 
                            new.pca.name = "pearsonpca_batch", 
                            new.umap.name = "pearsonumap_batch", 
                            var_features = NULL,
                            batch = "sample_id_col")

```


## Decontamination with decontX

``` {r decontX}

obj <- decontx_fun(obj, 
                  assay = "RNA", 
                  layer = "counts", 
                  labels = "pearson_leiden",
                  new.assay.name = "decontx")

```


## Resegmentation with FastReseg

``` {r fastreseg}

#Flagging cells for resegmentation
obj <- fastreseg_flag(obj, 
                      input_dir = COSMX_INPUTFILES_REPO, 
                      output_dir = COSMX_OUTPUTFILES_REPO, 
                      cluster_col = "pearson_leiden")


fastreseg_obj <- fastreseg_fullpipeline(obj, 
                                        input_dir = COSMX_INPUTFILES_REPO, 
                                        output_dir = COSMX_OUTPUTFILES_REPO, 
                                        cluster_col = "pearson_leiden")


```


## Idetifying markers with HieraType

```{r hieratype}

heatmap <- hieratype_fun(obj, 
                        assay = "RNA", 
                        cluster.name = "pearson_leiden", 
                        var_features = NULL)

hm_rna <- HieraType::add_ncells(HieraType::marker_heatmap(heatmap, topn = 20)) + 
  ggplot2::labs(title = "Heatmap")

```


## Tissue offsetting for despotX if analyzing multiple slides

I need to offset tissue coordinates for each FOV so that they do not overlap during despotX analysis.

```{r tissue_offsetting}

obj <- tissue_offset(obj)

```

## h5ad conversion for despotx

If you are working with multiple slides, ame sure you rum the tissue_offsetting function beforehand. 

``` {r h5ad_conversion}

# Wrappers for generating anndata object
# ======================================
python_anndata <- reticulate::import("anndata", convert = FALSE)
python_pandas <- reticulate::import("pandas", convert = FALSE)

# despotx's Python env (anndata 0.10.9) can't read the "nullable-string-array"
# encoding pandas>=3.0/anndata>=0.13 use by default for string dtypes; force
# plain object-dtype strings so the h5ad stays readable there.
python_pandas$set_option("future.infer_string", FALSE)
make_anndata <- function(X, obs, var, obsm = NULL) {
  obs_names <- rownames(X); if (is.null(obs_names)) obs_names <- rownames(obs)
  var_names <- colnames(X); if (is.null(var_names)) var_names <- rownames(var)
  if (is.data.frame(obs)) rownames(obs) <- obs_names
  if (is.data.frame(var)) rownames(var) <- var_names
  X <- anndata:::.check_matrix(X)   # same matrix-conversion helper the package itself uses
  ad <- python_anndata$AnnData(X = X, obs = obs, var = var, obsm = obsm)  # no dtype arg
  anndata::AnnDataR6$new(ad)         # wrap back into the normal R6 object ($write_h5ad(), $X, etc.)
}

# Conversion of Seurat object into a h5ad object / file
# =====================================================
tc = counts_mat <- Matrix::t(obj[["RNA"]]$counts)

# 2. Cell metadata — the column DeSpotX will condition on via --cell-type-col.
# sample_id/fov/cell_id are carried through so downstream reclustering
# (runScPearsonDimRed's batch correction) can run on despotx's Sherlock side
# without needing to re-derive them from the original Seurat object.
obs_df <- data.frame(
  cluster   = as.character(obj$pearson_leiden),
  cell_id   = as.character(obj$cell_ID),
  sample_id = as.character(obj$slidename),
  fov       = as.character(obj$fov),
  row.names = colnames(obj)
)

# 3. Gene metadata — var_names only strictly required by DeSpotX.
var_df <- data.frame(row.names = rownames(obj[["RNA"]]))

# 4. Spatial coordinates — must be EXACTLY 2D (x, y).
coords <- GetTissueCoordinates(obj, image = "all_slides")
coords <- coords[match(colnames(obj), coords$cell), c("x", "y")]
spatial_mat <- as.matrix(coords)

# 5. Assemble....
obj.adata <- make_anndata(
  X    = counts_mat,
  obs  = obs_df,
  var  = var_df,
  obsm = list(spatial = spatial_mat)
)

# 6. Force CSR before writing — via the raw Python object, not the R6 $X<- setter
py_ad <- obj.adata$.get_py_object()
sp <- reticulate::import("scipy.sparse", convert = FALSE)
py_ad$X <- sp$csr_matrix(py_ad$X)

# sanity check before writing
cat("X format:", as.character(py_ad$X$format), "\n")   # should print "csr"
cat("X shape:", paste(reticulate::py_to_r(py_ad$X$shape), collapse=" x "), "\n")

# 7. Write directly from the raw python object (bypasses the R6 wrapper entirely).
py_ad$write_h5ad(paste0(COSMX_OUTPUTFILES_REPO, date, "_output.h5ad"))

```

## Running despotX

This line code has to be run on terminal. 

```{r despotx}


despotx --h5ad /input_h5ad_path --cell-type-col cluster --spatial-key spatial --out /output_path/ --device cuda

```


## Import h5ad object
``` {r seurat_integration}
library(Seurat)
library(anndata)

h5ad_tma <- read_h5ad(Sys.getenv(h5ad_path))


# cells x genes (dgRMatrix) -> genes x cells (dgCMatrix), Seurat's expected orientation
X <- as(h5ad_tma$X, "CsparseMatrix")
rownames(X) <- rownames(h5ad_tma$obs)
colnames(X) <- rownames(h5ad_tma$var)
counts <- Matrix::t(X)

# Integrate into Seurat object
ft[["despotx"]] = Seurat::CreateAssayObject(counts)

saveRDS(object = ft_tma, file = paste0(COSMX_OUTPUTFILES_REPO, Sys.Date(), "_ft_wholeslides_despotx.rds"))

```


## Differential Gene Expression

```{r dge}

de_results <- lmm_fun(obj = obj,
                      assay = 'RNA', 
                      condition_col = "cancer_status", 
                      ref_condition = "early", 
                      var_condition = "advanced", 
                      donor_col = "slidename")
```


